Expanding tidy data principles to facilitate missing data exploration, visualization and assessment of imputations

### 2020-08-18 paper/review/response-to-review.Rmd

---
title: "Response to Review"
---

JSS 4108:

Nicholas John Tierney, Dianne Cook

Expanding tidy data principles to facilitate missing data exploration, visualization and assessment of imputations

---------------------------------------------------------
For further instructions on JSS style requirements please see the Style Files (in particular section 2.1 Style Checklist of jss.pdf) and FAQ at http://www.jstatsoft.org/about/submissions.

For further examples please see RECENT JSS papers.
---------------------------------------------------------

From the editorial team:

We apologize for the additional delays in processing 
your submission due to the COVID-19 pandemic.

The submission "Expanding tidy data principles to facilitate missing data
exploration, visualization and assessment of imputations" presents the R 
package naniar which provides data structures, summaries and visualizations 
for missing data.

Additional comments:

Please add some discussion on what kind of functionality is available in 
other statistical software packages than R in order to better embed the 
contribution not only within R, but the statistical software ecosystem in 
general.

Please provide a plain R script which when run reproduces all results 
presented in the manuscript. In general purling the Rmd file and editing 
the resulting R file  slightly in order to guide the user what each block 
does, removing knitr specific content,   etc. is suitable.

* Please add some discussion on what kind of functionality is available in 
other statistical software packages than R in order to better embed the 
contribution not only within R, but the statistical software ecosystem in 
general.

* Please provide a plain R script which when run reproduces all results 
presented in the manuscript. In general purling the Rmd file and editing 
the resulting R file slightly in order to guide the user what each block 
does, removing knitr specific content, etc. is suitable.

* The JSS style (https://www.jstatsoft.org/pages/view/style) should be 
followed more closely, especially using the title vs. sentence case, 
JSS-specific markup (\proglang, \pkg), and improving the BibTeX file. Some 
aspects of the style are handled incorrectly by "rticles", though (see
https://github.com/rstudio/rticles/issues/254).

* Note that the affiliation of the second author is misspelled.

Manuscript style comments:

o For the code layout in R publications, we typically distinguish input/output using Sinput/Soutput (or equivalently CodeInput/CodeOutput). Unless there are special reasons to format it differently, the input should use the text width (up to 76 or 77 characters) and be indented by two spaces, e.g.,

begin{Sinput}
R> example_model <- lm(response ~ variable1 + variable2 + variable3, 
+    weights = w, data = mydata)
\end{Sinput}

o If using "e.g." and "i.e." add a comma after the period to keep LaTeX from interpreting them as the end of a sentence, i.e.: "e.g., " and "i.e., ".

o The rule for capitalizing the starting letters of Figure, Section and Table is as follows: If you are referring to a particular figure/section/table then capitalize the first letter, otherwise use a lower-case first letter. For example, something shown in Section 4 vs. there are three sections in this paper.

o All captions should appear below the corresponding figure/table. The captions should be in sentence style and end with a period.  No additional formatting (such as \emph, f or \it) should be used for the caption.

o All table row/column headers should also be in sentence style. There should not be further footnote-style annotations in tables; these should all be placed in the caption.

o As a reminder, please make sure that:
	- \proglang, \pkg and \code have been used for highlighting throughout the paper (including titles and references), except where explicitly escaped.

References:

o Journal of the Royal Statistical Society B (not: Journal of the Royal Statistical Society, Series B)
	the EM Algorithm.” Journal of the Royal Statistical Society. Series B
the EM Algorithm.” Journal of the Royal Statistical Society. Series B
o Springer-Verlag (not: Springer)
	and GGobi. 1st edition. Springer Publishing Company, Incorporated.

o Please make sure that all software packages are \cite{}'d properly.
o All references should be in title style.
o See FAQ for specific reference instructions.

Code:

o Please make sure that the files needed to replicate all code/examples within the manuscript are included in a standalone replication script.

### 2020-09-04 paper/review/response-to-review.Rmd

The submission "Expanding tidy data principles to facilitate missing data
exploration, visualization and assessment of imputations" presents the R 
package naniar which provides data structures, summaries and visualizations 
for missing data.

Additional comments:

Please add some discussion on what kind of functionality is available in 
other statistical software packages than R in order to better embed the 
contribution not only within R, but the statistical software ecosystem in 
general.

> We have added the following:

- Missing data can be visualised in other software such as STATA, SAS, and SPSS
- They provide basic table frequencies
- SPSS provides visualisations on missingness patterns
- No software provides drop in functions to specifically visualise missing values.
- Although it is possible to visualise missing values after performing data manipulations to count and summarise the missingness.

@@

Please provide a plain R script which when run reproduces all results 
presented in the manuscript. In general purling the Rmd file and editing 
the resulting R file  slightly in order to guide the user what each block 
does, removing knitr specific content,   etc. is suitable.

> We have added a file, "tidy-missingdata-paper.R" that contains the R code from the paper

@@

Manuscript style comments:

o For the code layout in R publications, we typically distinguish input/output using Sinput/Soutput (or equivalently CodeInput/CodeOutput). Unless there are special reasons to format it differently, the input should use the text width (up to 76 or 77 characters) and be indented by two spaces, e.g.,

begin{Sinput}
R> example_model <- lm(response ~ variable1 + variable2 + variable3, 
+    weights = w, data = mydata)
\end{Sinput}

> I am not sure how to address this with `rticles`, I assume I'll need to update
to the latest version  and give that a go.

@@

o If using "e.g." and "i.e." add a comma after the period to keep LaTeX from interpreting them as the end of a sentence, i.e.: "e.g., " and "i.e., ".

> Instances of `e.g.` have been replaced with `e.g.,`. No cases of `i.e.` were found.

@@

o The rule for capitalizing the starting letters of Figure, Section and Table is as follows: If you are referring to a particular figure/section/table then capitalize the first letter, otherwise use a lower-case first letter. For example, something shown in Section 4 vs. there are three sections in this paper.

> Appropriate instances of `section` and `figure` have been replaced with `Section` and `Figure`.

@@

o All captions should appear below the corresponding figure/table. The captions should be in sentence style and end with a period.  No additional formatting (such as \emph, f or \it) should be used for the caption.

> I'm not sure it is possible to control the caption output of a kable Table, I will look into this.

### 2020-09-04 paper/tidy-missing-data-paper.Rmd
fix a compilation error and start adding responses to comments

---
documentclass: jss
author:
  - name: Nicholas Tierney
    affiliation: Monash University
    address: >
      Department of Econometrics and Business Statistics
      E774, Menzies Building
      Monash University, Clayton
    email: \email{nicholas.tierney@gmail.com}
    url: http://www.njtierney.com
  - name: Dianne Cook
    affiliation: Monsh University
title:
  formatted: "Expanding tidy data principles to facilitate missing data exploration, visualization and assessment of imputations"
  # If you use tex in the formatted title, also supply version without
  plain:     "Expanding tidy data principles to facilitate missing data exploration, visualization and assessment of imputations"
  # For running headers, if needed
  short:     "Explore missingness with \\pkg{naniar}"
abstract: >
  Despite the large body of research on missing value distributions and imputation, there is comparatively little literature with a focus on how to make it easy to handle, explore, and impute missing values in data. This paper addresses this gap. The new methodology builds upon tidy data principles, with the goal of integrating missing value handling as a key part of data analysis workflows. We define a new data structure, and a suite of new operations. Together, these provide a connected framework for handling, exploring, and imputing missing values. These methods are available in the R package `naniar`.
keywords:
  # at least one keyword must be supplied
  formatted: [statistical computing, statistical graphics, data science, data visualization, tidyverse, data pipeline, "\\proglang{R}"]
  plain:     [statistical computing, statistical graphics, data science, data visualization, tidyverse, data pipeline, R]
preamble: >
  \usepackage{amsmath}
  \usepackage[english]{babel}
output:
  bookdown::pdf_book:
    base_format: rticles::jss_article
    includes:
      in_header: header.tex
    keep_tex: yes
    citation_package: 'natbib'
bibliography: tidy-missing-data-paper.bib
---

@@

This paper is organized in the following way. The next section (\@ref(background)) provides the background to tidy data principles and tools (\@ref(tidy-data-concepts)) and missing data representations (\@ref(missing-data-rep-dep)). Section \@ref(existing-software) summarises existing software for handling missing data. The new work extending the tidy tools to facilitate exploring, visualizing, and imputing missing data is discussed in Section \@ref(extensions). Graphics (\@ref(graphics)) and Numerical summaries (\@ref(num-sum)) of missing values are then discussed. An application illustrating the use of the new methods is shown in Section \@ref(case-study). Section \@ref(discussion) discusses strengths, limitations, and future directions.

@@

Features of tidy data were formally described in @Wickham2014, and were discussed in terms of their importance for data science by @Donoho2017, and tools for data analysis. Tidy data is defined by @r4ds as:

> 1. Each variable must have its own column.
> 1. Each observation must have its own row.
> 1. Each value must have its own cell.

@@

There are many ways each value can be missing, we adopt the notation used in @VanBuuren2012. The information in $B$ can be used to arrive at three categories of missing values: Missing completely at random (MCAR), missing at random (MAR), and missing not at random (MNAR). The distribution of missing values in $b_{ij}$ can depend on the entire dataset, represented as $Y = (Y_{obs}, Y_{miss})$. This relationship can be defined by the _missing data model_ $Pr(b_{ij} | Y_{obs}, Y_{miss}, \psi)$, the probability of missingness is conditional on data observed, data missing, and some probability parameter of missingness, $\psi$. This helps to precisely define categories of missing values.

@@

<!-- (NOTE: Is the last sentence OK?) -->
<!-- (NOTE: trying to come up with a sentence that leads from this section to the next) -->

@@

The MANET (Missings Are Now Equally Treated) software @Unwin1996 focussed on multivariate categorical data, with missingness explicitly added as a category. MANET also provided univariate visualizations of missing data using linked brushing between a reference plot of the missingness for each variable, and a plot of the data as a histogram or barplot. The MANET software is no longer maintained and cannot be installed. The approach of @Swayne1998 in the software XGobi, further developed in ggobi [@Cook2007], focussed on multivariate quantitative data. Missingness is incorporated into plots in ggobi by setting them to be 10% below the minimum value.

@@

1. **Special missing values**: Building on **missing value labels**, the values in the shadow matrix can be "special" missing values, indicated by "NA_\<suffix\>". For example: `NA_instrument` uses a short label, "instrument", indicating instrument error resulting in missing values. These could be also used to indicate imputations, (Figure \@ref(fig:nabularfig)).

1. **Coordinated names**: Variable names in the shadow matrix gain a consistent short suffix, "_NA", keeping names coordinated throughout analysis (Figure \@ref(fig:nabularfig)). It makes a clear distinction with `var_NA` being a random variable of the missingness of a variable, `var`. This suffix is short and easy to remember during data analysis, and shifts the focus from the value of a variable, to its missingness state.

1. **Connectedness**: Binding the shadow matrix column-wise to the original data creates a single, connected, _nabular_ data format, in sync with the data. It is useful for visualization, summaries, and tracking imputed values, discussed in more detail in \@ref(nabular-data).

@@

_Nabular_ data binds the shadow matrix column-wise to the original data. It is a portmanteau of `NA` and `tabular`. _Nabular_ data explicitly links missing values to data, keeping corresponding observations together and removing the possibility of mismatching records (Figure \@ref(fig:nabularfig)). _Nabular_ data facilitates visualization and summaries by allowing the user to reference the missingness of a variable in a coordinated way: the missingness of `var`, as `var_NA` during analysis. _Nabular_ data is a snapshot of the missingness of the data. This means when _nabular_ data are imputed, those imputed values can easily be identified in analysis (`var` vs `var_NA`).  _Nabular_ data is not unlike classical data formats with quality or flag variables associated with each measured variable, e.g. Scripps CO2 data [@Keeling2005-scripps], GHCN data [@Durre2008-ghcn].

@@

Although a binary missingness matrix, $B$, could be generated during analysis instead of using a _nabular_ data structure, there are key advantages to using _nabular_ data. Firstly, missing value labels `NA` and `!NA` are clearer than TRUE or FALSE. Secondly, special missing values cannot be added easily during analysis with a logical matrix. Finally, the logical matrix cannot capture which values are imputed if imputation has already taken place. Imputing values on _nabular_ data automatically tracks these imputations.

Using additional columns to represent missingness information follows best practices for data organization, described in @Ellis2017 and @Broman2017: (1) Keep one thing in a cell and (2) Describe additional features of variables in a second column. Here they suggest to indicate censored data with an extra variable called "VariableNameCensored", which would be TRUE if censored, otherwise FALSE. This information can now be represented in the shadow columns as special missing values. Encoding special missing values is achieved by defining logical conditions and suffixes.  This is implemented with the `recode_shadow` function in \pkg{naniar} (\@ref(verbs-recode)).

@@

Understanding the missingness structure can be improved by adding summary information alongside the data. For example, in @Tierney2015, the proportion of
missings in a row is used as the outcome in a model to identify variables
important in predicting missingness structures. `naniar` implements a series of functions to add these missingness summaries to the data, starting with `add_`. These are inspired by the `add_count` function in `dplyr`, which adds count information for specified groups or conditions. `naniar` provides operations to add the number or proportion of missingness, the missingness cluster, or the presence of any missings, with: `add_n_miss()`, `add_prop_miss()`, `add_miss_cluster`, and `add_any_miss()`, respectively (Table \@ref(tab:add-missing-info)). There are also functions for adding information about shadow values, and readable labels for any missing values with `add_label_shadow()` and `add_label_missings()`.

@@

_Nabular_ data has the shadow matrix column-bound to existing data. This facilitates visualization and summaries, and allows for imputed values to be tracked. _Nabular_ data can be created with `nabular()`:

@@

Unusual or spurious data values are often identified and `flagged`. For example, there might be special codes to mark an individual dropping out of a study, known instrument failure in weather instruments, or for values censored in analysis. `naniar` provides tools to encode these special types of missingness in the shadow matrix of  _nabular_  data, using `recode_shadow()`. This requires specifying the variable to contain the flagged value, the condition for flagging, and a suffix. This is then recoded as a new factor level in the shadow matrix, so every column is aware of all possible new values of missingness. For example, -99 could be recoded to indicate a broken machine sensor for the variable `x` with the following:

@@

The  first step is to get an overview of the extent of the missingness. Overview visualizations for variables and cases are provided with `gg_miss_var` and `gg_miss_case` (Figure \@ref(fig:gg-miss-case-var)A), drawing attention to the amount of missings, and ordering by missingness. The "airquality" dataset (from base R), is shown, and contains daily air quality measurements in New York, from May to September, 1973. We learn from figure \@ref(fig:gg-miss-case-var)A-B, that two variables contain missings, approximately one third of observations have one missing value, and a tiny number of observations are missing across two variables. These overview plots are created from the shadow matrix in long form (Figure \@ref(fig:nabularfig)). Numerical statistics can also be reported (Section \@ref(num-sum)).

@@

Missing values are typically not shown for univariate visualizations such as histograms or densities. Two ways to use _nabular_ data to present univariate data with missings are discussed. The first imputes values below the range to facilitate visualizations. The second displays two plots of the same variable according to the missingness of a chosen variable.

**Imputing values below the range**. To visualize the amount of missings in each variable, the data is transformed into _nabular_ form, then values are imputed below the range of data using `impute_below_all`. Figure \@ref(fig:impute-shift-histogram)A shows a histogram of Ozone values on the right in green, and the histogram of missing ozone values on the left, in orange. The missings in Ozone are imputed and `Ozone_NA` is  mapped to the fill aesthetic. Nabular data facilitates adding counts of missingness to a histogram, allowing examination of a variables' distribution of values, and also the magnitude of missings.

@@

To visualize missing values in two dimensions the missing values can be placed in plot margins, by imputing values below the range of the data.  Using _nabular_ data identifies imputed values, and color makes missingness pre-attentive [@treisman1985]. The steps of imputing and coloring are combined into `geom_miss_point()`. Figure \@ref(fig:geom-miss) shows a mostly uniform spread of missing values for Solar.R and Ozone. As `geom_miss_point()` is a defined \pkg{ggplot2} geometry, it works with features such as faceting and mapping other variables to graphical aesthetics.

@@

It is useful to explore summaries and tabulations within groups of a dataset.  `naniar` works with `dplyr`'s `group_by` operator to produce grouped summaries, which work well with the "pipe" operator. The code and table below \@ref(tab:group-miss-var-summary) show an example of missing data summaries for airquality, grouped by month.

@@

dat_house_cc <- na.omit(dat_house)

@@

The _nabular_ data structures discussed in the paper are simple by design, to promote flexibility.  They could be used to create different visualizations than were shown in the paper. The analyst can use the data structures to decide on appropriate visualization for their problem. The data structures could also be used to support interactive graphics, in the manner of MANET and ggobi. Linking the plots (via linked brushing) could facilitate exploration of missingness, and could be implemented with \pkg{plotly} [@plotly] for added interactivity. Animating between different sets of imputed values could also be possible with packages like \pkg{gganimate} [@gganimate].

Other data structures such as spatial data, time series, networks, and
longitudinal data would be supported by the inherently tabular,
_nabular_ data, if they are first structured as wide tidy format. Large data may need special handling, and additional features like efficient storage of purely imputed values and lazy evaluation. Special missing value codes could be improved by creating special classes, or expanding low level representation of `NA` at the source code level.

The methodology described in this paper can be used in conjunction with other approaches to understand multivariate missingness dependencies (e.g. decision trees [@Tierney2015], latent group analysis [@Barnett2017], and PCA [@FactoMineR]). Evaluating imputed values using a testing framework like @VanBuuren2012 is also supported.

>>>>

---
documentclass: jss
author:
  - name: Nicholas Tierney
    affiliation: Monash University
    address: >
      Department of Econometrics and Business Statistics
      E774, Menzies Building
      Monash University, Clayton
    email: \email{nicholas.tierney@gmail.com}
    url: http://www.njtierney.com
  - name: Dianne Cook
    affiliation: Monash University
title:
  formatted: "Expanding tidy data principles to facilitate missing data exploration, visualization and assessment of imputations"
  # If you use tex in the formatted title, also supply version without
  plain:     "Expanding tidy data principles to facilitate missing data exploration, visualization and assessment of imputations"
  # For running headers, if needed
  short:     "Explore missingness with \\pkg{naniar}"
abstract: >
  Despite the large body of research on missing value distributions and imputation, there is comparatively little literature with a focus on how to make it easy to handle, explore, and impute missing values in data. This paper addresses this gap. The new methodology builds upon tidy data principles, with the goal of integrating missing value handling as a key part of data analysis workflows. We define a new data structure, and a suite of new operations. Together, these provide a connected framework for handling, exploring, and imputing missing values. These methods are available in the R package `naniar`.
keywords:
  # at least one keyword must be supplied
  formatted: [statistical computing, statistical graphics, data science, data visualization, tidyverse, data pipeline, "\\proglang{R}"]
  plain:     [statistical computing, statistical graphics, data science, data visualization, tidyverse, data pipeline, R]
preamble: >
  \usepackage{amsmath}
  \usepackage[english]{babel}
output:
  bookdown::pdf_book:
    base_format: rticles::jss_article
    includes:
      in_header: header.tex
    keep_tex: yes
    citation_package: 'natbib'
bibliography: tidy-missing-data-paper.bib
---

@@

This paper is organized in the following way. Section (\@ref(background)) provides the background to tidy data principles and tools (\@ref(tidy-data-concepts)) and missing data representations (\@ref(missing-data-rep-dep)). Section \@ref(existing-software) summarises existing software for handling missing data. The new work extending the tidy tools to facilitate exploring, visualizing, and imputing missing data is discussed in Section \@ref(extensions). Graphics (\@ref(graphics)) and Numerical summaries (\@ref(num-sum)) of missing values are then discussed. An application illustrating the use of the new methods is shown in Section \@ref(case-study). Section \@ref(discussion) discusses strengths, limitations, and future directions.

@@

Features of tidy data were formally described in [@Wickham2014], and were discussed in terms of their importance for data science by [@Donoho2017], and tools for data analysis. Tidy data is defined by [@r4ds] as:

> 1.  Each variable must have its own column.
> 2.  Each observation must have its own row.
> 3.  Each value must have its own cell.

@@

There are many ways each value can be missing, we adopt the notation used in [@VanBuuren2012]. The information in $B$ can be used to arrive at three categories of missing values: Missing completely at random (MCAR), missing at random (MAR), and missing not at random (MNAR). The distribution of missing values in $b_{ij}$ can depend on the entire dataset, represented as $Y = (Y_{obs}, Y_{miss})$. This relationship can be defined by the *missing data model* $Pr(b_{ij} | Y_{obs}, Y_{miss}, \psi)$, the probability of missingness is conditional on data observed, data missing, and some probability parameter of missingness, $\psi$. This helps to precisely define categories of missing values.

@@

<!-- (NOTE: Is the last sentence OK?) -->

<!-- (NOTE: trying to come up with a sentence that leads from this section to the next) -->

@@

The MANET (Missings Are Now Equally Treated) software [@Unwin1996] focussed on multivariate categorical data, with missingness explicitly added as a category. MANET also provided univariate visualizations of missing data using linked brushing between a reference plot of the missingness for each variable, and a plot of the data as a histogram or barplot. The MANET software is no longer maintained and cannot be installed. The approach of [@Swayne1998] in the software XGobi, further developed in ggobi [@Cook2007], focussed on multivariate quantitative data. Missingness is incorporated into plots in ggobi by setting them to be 10% below the minimum value.

@@

2.  **Special missing values**: Building on **missing value labels**, the values in the shadow matrix can be "special" missing values, indicated by "NA\_\<suffix\>". For example: `NA_instrument` uses a short label, "instrument", indicating instrument error resulting in missing values. These could be also used to indicate imputations, (Figure \@ref(fig:nabularfig)).

3.  **Coordinated names**: Variable names in the shadow matrix gain a consistent short suffix, \"\_NA\", keeping names coordinated throughout analysis (Figure \@ref(fig:nabularfig)). It makes a clear distinction with `var_NA` being a random variable of the missingness of a variable, `var`. This suffix is short and easy to remember during data analysis, and shifts the focus from the value of a variable, to its missingness state.

4.  **Connectedness**: Binding the shadow matrix column-wise to the original data creates a single, connected, *nabular* data format, in sync with the data. It is useful for visualization, summaries, and tracking imputed values, discussed in more detail in \@ref(nabular-data).

@@

*Nabular* data binds the shadow matrix column-wise to the original data. It is a portmanteau of `NA` and `tabular`. *Nabular* data explicitly links missing values to data, keeping corresponding observations together and removing the possibility of mismatching records (Figure \@ref(fig:nabularfig)). *Nabular* data facilitates visualization and summaries by allowing the user to reference the missingness of a variable in a coordinated way: the missingness of `var`, as `var_NA` during analysis. *Nabular* data is a snapshot of the missingness of the data. This means when *nabular* data are imputed, those imputed values can easily be identified in analysis (`var` vs `var_NA`). *Nabular* data is not unlike classical data formats with quality or flag variables associated with each measured variable, e.g., Scripps CO2 data [@Keeling2005-scripps], GHCN data [@Durre2008-ghcn].

@@

Although a binary missingness matrix, $B$, could be generated during analysis instead of using a *nabular* data structure, there are key advantages to using *nabular* data. Firstly, missing value labels `NA` and `!NA` are clearer than TRUE or FALSE. Secondly, special missing values cannot be added easily during analysis with a logical matrix. Finally, the logical matrix cannot capture which values are imputed if imputation has already taken place. Imputing values on *nabular* data automatically tracks these imputations.

Using additional columns to represent missingness information follows best practices for data organization, described in [@Ellis2017] and [@Broman2017]: (1) Keep one thing in a cell and (2) Describe additional features of variables in a second column. Here they suggest to indicate censored data with an extra variable called "VariableNameCensored", which would be TRUE if censored, otherwise FALSE. This information can now be represented in the shadow columns as special missing values. Encoding special missing values is achieved by defining logical conditions and suffixes. This is implemented with the `recode_shadow` function in \pkg{naniar} (\@ref(verbs-recode)).

@@

Understanding the missingness structure can be improved by adding summary information alongside the data. For example, in [@Tierney2015], the proportion of missings in a row is used as the outcome in a model to identify variables important in predicting missingness structures. `naniar` implements a series of functions to add these missingness summaries to the data, starting with `add_`. These are inspired by the `add_count` function in `dplyr`, which adds count information for specified groups or conditions. `naniar` provides operations to add the number or proportion of missingness, the missingness cluster, or the presence of any missings, with: `add_n_miss()`, `add_prop_miss()`, `add_miss_cluster`, and `add_any_miss()`, respectively (Table \@ref(tab:add-missing-info)). There are also functions for adding information about shadow values, and readable labels for any missing values with `add_label_shadow()` and `add_label_missings()`.

@@

*Nabular* data has the shadow matrix column-bound to existing data. This facilitates visualization and summaries, and allows for imputed values to be tracked. *Nabular* data can be created with `nabular()`:

@@

Unusual or spurious data values are often identified and `flagged`. For example, there might be special codes to mark an individual dropping out of a study, known instrument failure in weather instruments, or for values censored in analysis. `naniar` provides tools to encode these special types of missingness in the shadow matrix of *nabular* data, using `recode_shadow()`. This requires specifying the variable to contain the flagged value, the condition for flagging, and a suffix. This is then recoded as a new factor level in the shadow matrix, so every column is aware of all possible new values of missingness. For example, -99 could be recoded to indicate a broken machine sensor for the variable `x` with the following:

@@

The first step is to get an overview of the extent of the missingness. Overview visualizations for variables and cases are provided with `gg_miss_var` and `gg_miss_case` (Figure \@ref(fig:gg-miss-case-var)A), drawing attention to the amount of missings, and ordering by missingness. The "airquality" dataset (from base R), is shown, and contains daily air quality measurements in New York, from May to September, 1973. We learn from Figure \@ref(fig:gg-miss-case-var)A-B, that two variables contain missings, approximately one third of observations have one missing value, and a tiny number of observations are missing across two variables. These overview plots are created from the shadow matrix in long form (Figure \@ref(fig:nabularfig)). Numerical statistics can also be reported (Section \@ref(num-sum)).

@@

Missing values are typically not shown for univariate visualizations such as histograms or densities. Two ways to use *nabular* data to present univariate data with missings are discussed. The first imputes values below the range to facilitate visualizations. The second displays two plots of the same variable according to the missingness of a chosen variable.

**Imputing values below the range**. To visualize the amount of missings in each variable, the data is transformed into *nabular* form, then values are imputed below the range of data using `impute_below_all`. Figure \@ref(fig:impute-shift-histogram)A shows a histogram of Ozone values on the right in green, and the histogram of missing ozone values on the left, in orange. The missings in Ozone are imputed and `Ozone_NA` is mapped to the fill aesthetic. Nabular data facilitates adding counts of missingness to a histogram, allowing examination of a variables' distribution of values, and also the magnitude of missings.

@@

To visualize missing values in two dimensions the missing values can be placed in plot margins, by imputing values below the range of the data. Using *nabular* data identifies imputed values, and color makes missingness pre-attentive [@treisman1985]. The steps of imputing and coloring are combined into `geom_miss_point()`. Figure \@ref(fig:geom-miss) shows a mostly uniform spread of missing values for Solar.R and Ozone. As `geom_miss_point()` is a defined \pkg{ggplot2} geometry, it works with features such as faceting and mapping other variables to graphical aesthetics.

@@

It is useful to explore summaries and tabulations within groups of a dataset. `naniar` works with `dplyr`'s `group_by` operator to produce grouped summaries, which work well with the "pipe" operator. The code and Table \@ref(tab:group-miss-var-summary) below show an example of missing data summaries for airquality, grouped by month.

@@

dat_house_cc <- dat_house %>% nabular() %>% add_label_shadow() %>% na.omit()

@@

The *nabular* data structures discussed in the paper are simple by design, to promote flexibility. They could be used to create different visualizations than were shown in the paper. The analyst can use the data structures to decide on appropriate visualization for their problem. The data structures could also be used to support interactive graphics, in the manner of MANET and ggobi. Linking the plots (via linked brushing) could facilitate exploration of missingness, and could be implemented with \pkg{plotly} [@plotly] for added interactivity. Animating between different sets of imputed values could also be possible with packages like \pkg{gganimate} [@gganimate].

Other data structures such as spatial data, time series, networks, and longitudinal data would be supported by the inherently tabular, *nabular* data, if they are first structured as wide tidy format. Large data may need special handling, and additional features like efficient storage of purely imputed values and lazy evaluation. Special missing value codes could be improved by creating special classes, or expanding low level representation of `NA` at the source code level.

The methodology described in this paper can be used in conjunction with other approaches to understand multivariate missingness dependencies (e.g., decision trees [@Tierney2015], latent group analysis [@Barnett2017], and PCA [@FactoMineR]). Evaluating imputed values using a testing framework like [@VanBuuren2012] is also supported.
